---
title: "density_gaza"
author: "Olaf König"
format: html
editor: visual
---

# Objective

-   Density with GHS Data for Gaza strip

-   Density with GHS Geneva and Schaffhausen

-   calculate area approximation of territory which is not militarized nor under displacement order

-   calculate current population density, if the whole population lives in these parts...

-   check the results

------------------------------------------------------------------------

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(mapview)
```

## Directories path

```{r}

dir_input <- "data_input/data_raw"
dir_proccesed <- "data_input/data_processed"
```

## Import dataset

### Global Human Settlement Layer

```{r}

ghs <- rast("/Users/bajj@mediait.ch/Documents/LEDE_Program/Projets/data/population/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0.tif")
```

## Vector geometries

```{r}

sf_swiss <- read_sf("data_input/data_raw/Cantons_G1_20250101_2056.geojson") %>% st_transform(4326)

sf_geneva <- sf_swiss %>%
  filter(KTKZ == "GE") %>% mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )

sf_schaffhausen <- sf_swiss %>%
  filter(KTKZ == "SH") %>% mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )

uno_gaza_strip <- read_sf(file.path(dir_proccesed, "uno_gaza_strip_level1.gpkg")) %>%
  mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )




```

Export geometries

```{r}

# Exporter les géométries
sf_geneva %>% 
  st_write("data_input/data_processed/geneva.geojson", delete_dsn = TRUE)

sf_schaffhausen %>% 
  st_write("data_input/data_processed/schaffhausen.geojson", delete_dsn = TRUE)
```

### Militarized and displacement zone, and zone without restrictions

```{r}

ocha_miliarized_zone <- read_sf("data_input/data_processed/mathias_ocha/militarized_zone.gpkg") %>% 
  mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )

ocha_displacement_order <- read_sf("data_input/data_processed/mathias_ocha/displacement_orders.gpkg") %>% 
  mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )

uno_gaza_strip_no_restriction <- read_sf("data_input/data_processed/mathias_ocha/gaza_minus_militarized/gaza_minus_merged_cleaned.geojson") %>%
  mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  )
```

## Crop GHS layer

```{r}

ghs_gaza_strip <- crop(ghs, uno_gaza_strip, mask = TRUE)

ghs_geneva <- crop(ghs, sf_geneva, mask = TRUE)

ghs_schaffhausen <- crop(ghs, sf_schaffhausen, mask = TRUE)

ghs_gaza_no_restriction <- crop(ghs, uno_gaza_strip_no_restriction, 
                                mask = TRUE)

# ghs_gaza_strip %>%
#   terra::writeRaster(file.path(dir_proccesed, "ghs_gaza_strip.tif"))
```

```{r}

ghs_gaza_strip %>%
  terra::writeRaster(file.path(dir_proccesed, "ghs_gaza_strip.tif"))

ghs_geneva %>%
  terra::writeRaster(file.path(dir_proccesed, "ghs_geneva.tif"))

ghs_schaffhausen %>%
  terra::writeRaster(file.path(dir_proccesed, "ghs_schaffhausen.tif"))
```

## Population calculation with cropped GHS

```{r}


# Calcul de la somme pour chaque raster
sum_geneva <- global(ghs_geneva, "sum", na.rm = TRUE)
sum_schaffhausen <- global(ghs_schaffhausen, "sum", na.rm = TRUE)
sum_gaza_strip <- global(ghs_gaza_strip, "sum", na.rm = TRUE)
sum_gaza_no_restriction <- global(ghs_gaza_no_restriction, "sum", na.rm = TRUE)

# # Affichage des résultats
# print(sum_geneva)
# print(sum_schaffhausen)
# print(sum_gaza_strip)
# print(sum_gaza_no_restriction)
```

```{r}

# Création d'un dataframe avec les résultats
resultats <- data.frame(
  territoire = c("Geneva", "Schaffhausen", "Gaza Strip", "Gaza No Restriction"),
  somme_ghs = c(
    global(ghs_geneva, "sum", na.rm = TRUE)$sum,
    global(ghs_schaffhausen, "sum", na.rm = TRUE)$sum,
    global(ghs_gaza_strip, "sum", na.rm = TRUE)$sum,
    global(ghs_gaza_no_restriction, "sum", na.rm = TRUE)$sum
  )
)

print(resultats)
```

## Gaza strip no restriction - simulation of 2.1 mio habitants in this area

```{r}

# Source: https://www.ochaopt.org/content/reported-impact-snapshot-gaza-strip-13-august-2025

population_current <- 2100000

# On charge les immeubles
osm_buildings_unosat_damage <- read_sf(file.path(dir_proccesed,
                     "osm_buildings_unosat_damage_greenhouses_excluded.gpkg"))

# On ajoute un buffer de 5 mètres
# Pour améliorer les performances sur 327k objets
osm_buildings_buffered <- osm_buildings_unosat_damage %>%
  st_transform(32636) %>%           # UTM 36N
  st_buffer(dist = 5) %>%           # Buffer 5m
  st_transform(4326)                # Retour WGS84





# On enlève les immeubles des batiments bufferisés
ghs_gaza_no_restriction_wihtout_houses <- st_difference(uno_gaza_strip_no_restriction, 
                          st_union(osm_buildings_buffered))

  
  
  

```
