---
title: "density_gaza"
author: "Olaf König"
format: html
editor: visual
---

# Objective

-   Density with GHS Data for Gaza strip

-   Density with GHS Geneva and Schaffhausen

-   calculate area approximation of territory which is not militarized nor under displacement order

-   calculate current population density, if the whole population lives in these parts...

-   check the results

------------------------------------------------------------------------

```{r setup}

library(tidyverse)
library(terra)
library(sf)
library(mapview)
```

## Directories path

```{r}

dir_input <- "data_input/data_raw"
dir_proccesed <- "data_input/data_processed"
```

## Import dataset

### Global Human Settlement Layer

```{r}

library(terra)

ghs <- rast("/Users/bajj@mediait.ch/Documents/LEDE_Program/Projets/data/population/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0.tif")
```

## Vector geometries

```{r}

sf_swiss <- read_sf("data_input/data_raw/Cantons_G1_20250101_2056.geojson") %>% st_transform(4326)

sf_geneva <- sf_swiss %>%
  filter(KTKZ == "GE") %>% 
  mutate(area = as.numeric(st_area(.)))

sf_schaffhausen <- sf_swiss %>%
  filter(KTKZ == "SH") %>% 
  st_area()

uno_gaza_strip <- read_sf(file.path(dir_proccesed, "uno_gaza_strip_level1.gpkg")) %>%  
  summarise() %>% 
  st_area()
```

## Crop GHS layer

```{r}

ghs_gaza_strip <- crop(ghs, uno_gaza_strip, mask = TRUE)

# ghs_gaza_strip %>%
#   terra::writeRaster(file.path(dir_proccesed, "ghs_gaza_strip.tif"))
```
