---
title: "get_geodata"
author: "Olaf König"
format: html
editor: visual
---

## Gaza stripes geodata access

## Objective

Get geodata from UNOSAT Damage Assessment to create maps under:

Geodata: <https://unosat.org/products/4165>

The GDB was opened in QGIS, and then exported as a GPKG file.

Metadata here: <https://data.humdata.org/dataset/unosat-gaza-strip-comprehensive-damage-assessment-september-2024>

Get buildings from OSM via Geofabrik

<https://download.geofabrik.de/asia/israel-and-palestine.html>

Get GHS density population from Copernicus 2025

<https://human-settlement.emergency.copernicus.eu/ghs_pop2023.php>

Exports:

-   crossed_osm_building with unosat data

-   ghs gaza strip

-   boundaries gaza level1 et level2

-   boundaries region (for inset map)

Compute share of destruction for non-restriction zone

------------------------------------------------------------------------

```{r setup}

library(tidyverse)
library(tidylog)
library(sf)
library(mapview)
library(rnaturalearth)
library(rnaturalearthdata)
```

```{r}

dir_input <- "data_input/data_raw"
dir_proccesed <- "data_input/data_processed"
```

# Load files

## UNOSAT file

```{r}

sf_unosat <- read_sf(file.path(dir_input, "unosat_cda.gpkg"))

# glimpse(sf_unosat)
```

## OSM building file (Israel and Palestin as whole)

```{r}

osm_buildings <- read_sf(file.path(dir_input, "israel-and-palestine-latest-free.shp/gis_osm_buildings_a_free_1.shp"))
```

## Combining both

```{r}

osm_buildings_unosat_damage <- st_join(osm_buildings, sf_unosat)

```

## Get Natural earth boundaries

```{r}

nr_countries <- ne_countries(scale = "medium", country =  c(
  "Israel", "Palestine", "Jordan", "Egypt", "Syria",
  "Lebanon", "Iraq", "Saudi Arabia", "Turkey", "Cyprus"))
```

```{r}

nr_countries %>%
  write_sf(file.path(dir_proccesed, "nr_countries_inset_map.gpkg"))
```

## Get Palestine boundaries

Invalid date in "validTo"

```{r}

uno_palestine_level2 <- read_sf("data_input/data_raw/pse_adm_pamop_20231019_shp/pse_admbnda_adm2_pamop_20231019.shp")

uno_gaza_strip_level2 <- uno_palestine_level2 %>% 
    filter(ADM1_EN == "Gaza Strip")

uno_gaza_strip_level2 %>% 
  select(-validTo) %>% 
  write_sf(file.path(dir_proccesed, "uno_gaza_strip_level2.gpkg"))

uno_gaza_strip <- uno_palestine_level2 %>% 
  filter(ADM1_EN == "Gaza Strip")

uno_gaza_strip %>% 
    select(-validTo) %>% 
    write_sf(file.path(dir_proccesed, "uno_gaza_strip_level1.gpkg"))


uno_gaza_strip_no_restriction <- read_sf("data_input/data_processed/mathias_ocha/gaza_minus_militarized/gaza_no_restriction_clean.gpkg") %>%
  mutate(
    area_m2  = as.numeric(st_area(.)),           # surface en m²
    area_km2 = round(area_m2 / 1e6, 2)           # surface en km², arrondi à 2 décimales
  ) %>% filter(!st_is_empty(geom))
```

## Filter osm building

```{r}

osm_buildings_unosat_damage <- osm_buildings_unosat_damage %>%
  st_filter(uno_gaza_strip)

osm_buildings_unosat_damage %>%
  write_sf(file.path(dir_proccesed, "osm_buildings_unosat_damage.gpkg"))

osm_buildings_unosat_damage %>%
  filter(is.na(type) | type != "greenhouse") %>%
  write_sf(file.path(dir_proccesed,
                     "osm_buildings_unosat_damage_greenhouses_excluded.gpkg"))

```

### Check of OSM building groups (greenhouse, etc.)

```{r}
# Tableau simple avec n
synthese <- osm_buildings_unosat_damage %>% 
  st_drop_geometry() %>% 
  count(type, name = "n")

# Avec pourcentages en plus
synthese <- osm_buildings_unosat_damage |>
  count(fclass, name = "n") |>
  mutate(
    pourcentage = round(n / sum(n) * 100, 1),
    pourcentage_label = paste0(pourcentage, "%")
  )

```

```{r}

# mapview(osm_buildings_unosat_damage %>% 
#           filter(type == "greenhouse"), 
#         alpha.regions = 0.2, 
#         col.regions = "red")
```

## Population density from GHS Copernicus

```{r}
library(terra)

ghs <- rast("/Users/bajj@mediait.ch/Documents/LEDE_Program/Projets/data/population/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0/GHS_POP_E2025_GLOBE_R2023A_4326_3ss_V1_0.tif")

ghs_gaza_strip <- crop(ghs, uno_gaza_strip, mask = TRUE)

ghs_gaza_strip %>%
  terra::writeRaster(file.path(dir_proccesed, "ghs_gaza_strip.tif"))
```

# Process

### How much destruction in no-restriction zone?

```{r}
# Filter osm buildings in the no restriction zone
osm_buildings_unosat_damage_no_restricion_zone <- osm_buildings_unosat_damage %>%
  st_filter(uno_gaza_strip_no_restriction)

# check the results
sampled <- osm_buildings_unosat_damage_no_restricion_zone |>
  dplyr::slice_sample(n = 1000) |>
  dplyr::mutate(Main_Damage_Site_Class_13 = as.factor(Main_Damage_Site_Class_13))

mapview(uno_gaza_strip_no_restriction) +
  mapview(sampled, zcol = "Main_Damage_Site_Class_13")
# Si tu préfères des points: mapview(st_centroid(sampled), zcol = "Main_Damage_Site_Class_13")
```

```{r}

# How much is left?
df_damage_class_no_restriction_zone <- sf_unosat %>% 
  st_drop_geometry() %>% 
  group_by(Main_Damage_Site_Class_13) %>% 
  summarise(
    n = n()
  )
```

### Agregated by **Damage Class**

```{r}

df_damage_class_no_restriction_zone <- osm_buildings_unosat_damage_no_restricion_zone %>% 
  st_drop_geometry() %>% 
  group_by(Main_Damage_Site_Class_13) %>% 
  summarise(n = n()) %>% 
  mutate(
    pct = n / sum(n) * 100
  )

print(df_damage_class_no_restriction_zone)
```

```{r}

df_damage_class %>%  
  ggplot(aes(x = n, 
             y = reorder(Main_Damage_Site_Class_13, n))) +
  geom_col(fill = "steelblue") +
  labs(
    x = "n",
    y = "Damage class"
  ) +
  theme_minimal()
```

# 
